---
title: "Stat 558: Project 1"
author: "Bryan Sandor"
format: html
editor_options: 
  chunk_output_type: console
---

# Preamble
The goal is to write functions manipulating and processing data sets originating from a census application programming interface (API). _Generic_ functions will be created to automatically summarize and plot some returned data. We load the necessary libraries in this first code section.

```{r necessary_packages}
library(tidyverse)
library(dplyr)
library(httr)
library(jsonlite)
library(janitor)
```


## Data
We will be using the Public Use Microdata Sample (PUMS) Census API. Information about variables from selected years may be found [here](https://api.census.gov/data/2024/acs/acs1/pums/variables.html) by changing the year section (here, $2024$).

# Data Processing
1. Here is an example call for Tennessee (code `STATE = 47`) from $2024$ citizens having a doctorate (code `SCHL = 24`), that returns the categorical variables `SEX` and marital status (`MAR`) and the numeric variable weight (`PWGTP`).

```{r data_processing_bullet_1}
census <- GET("https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:47&SCHL=24&key=5f3b8018114ef7902b34be730306e5ebf9102875")
```

2. Then we use _helper_ functions to "clean" up the data.

```{r data_processing_bullet_2}
census_tibble <- census$content |>
  rawToChar() |>
  fromJSON() |>
  as_tibble(.name_repair = "minimal") |>
  row_to_names(row_number = 1) # "promotes" first row to column headers

census_tibble
```

3. Now we write a function 

```{r data_processing_bullet_3}
census_query <- function(year = 2022,
                         numvar = c("AGEP", "PWGTP"),
                         catvar = "SEX",
                         geolevel = "STATE",
                         subset = NULL){
  
  if (year < 2010 || year == 2020 || year > 2024){
    print("Please enter a year between 2010 and 2024, inclusive; 2020 is also not available.")
  }
  
  if (length(numvar) < 2 || 
      any(!(numvar %in% c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP",
                             "JWMNP", "PWGTP")))){
    print("Please enter at least *two* numerical values from 'AGEP', 'GASP', 'GRPIP', 'JWAP', 'JWDP', or 'JWMNP'.")
  }
  
  if (any(!(catvar %in% c("FER", "HHL", "SCH", "SCHL", "SEX")))){
    print("Please enter at least one categorical values from 'FER', 'HHL', 'SCH', 'SCHL', or 'SEX' (default).")
  }
  
  c(numvar, catvar)
}
```

