---
title: "Stat 558: Project 1"
author: "Bryan Sandor"
format: html
editor_options: 
  chunk_output_type: console
---

# Preamble
The goal is to write functions manipulating and processing data sets originating from a census application programming interface (API). _Generic_ functions will be created to automatically summarize and plot some returned data. We load the necessary libraries in this first code section.

```{r necessary_packages}
library(tidyverse)
library(dplyr)
library(httr)
library(jsonlite)
library(janitor)
library(lubridate)
```


## Data
We will be using the Public Use Microdata Sample (PUMS) Census API. Information about variables from selected years may be found [here](https://api.census.gov/data/2024/acs/acs1/pums/variables.html) by changing the year section (here, $2024$).

# Data Processing
1. Here is an example call for Tennessee (code `STATE = 47`) from $2024$ citizens having a doctorate (code `SCHL = 24`), that returns the categorical variables `SEX` and marital status (`MAR`) and the numeric variable weight (`PWGTP`).

```{r data_processing_bullet_1}
census_tn <- GET("https://api.census.gov/data/2024/acs/acs1/pums?get=SEX,PWGTP,MAR&for=state:47&SCHL=24&key=5f3b8018114ef7902b34be730306e5ebf9102875")
```

2. Then we use _helper_ functions to "clean" up the data.

```{r data_processing_bullet_2}
census_tn_tibble <- census_tn$content |>
  rawToChar() |>
  fromJSON() |>
  as_tibble(.name_repair = "minimal") |>
  row_to_names(row_number = 1) # "promotes" first row to column headers

census_tn_tibble
```

3. Now we write a function 

```{r data_processing_bullet_3}
census_query <- function(year = 2022,
                         numvar = c("AGEP", "PWGTP"),
                         catvar = "SEX",
                         geolevel = "STATE",
                         subset = NULL){

# geographic level dictionaries
states <- list("AK" = "02", "AL" = "01", "AR" = "05", "AZ" = "04", "CA" = "06", "CO" = "08", "CT" = "09", "DC" = "11", "DE" = "10", "FL" = "12", "GA" = "13", "HI" = "15", "IA" = "19", "ID" = "16", "IL" = "17", "IN" = "18", "KS" = "20", "KY" = "21", "LA" = "22", "MA" = "25", "MD" = "24", "ME" = "23", "MI" = "26", "MN" = "27", "MO" = "29", "MS" = "28", "MT" = "30", "NC" = "37", "ND" = "38", "NE" = "31", "NH" = "33", "NJ" = "34", "NM" = "35", "NV" = "32", "NY" = "36", "OH" = "39", "OK" = "40", "OR" = "41", "PA" = "42", "RI" = "44", "SC" = "45", "SD" = "46", "TN" = "47", "TX" = "48", "UT" = "49", "VA" = "51", "VT" = "50", "WA" = "53", "WI" = "55", "WV" = "54", "WY" = "56")

regions <- list("NORTHEAST" = '1', "MIDWEST" = '2', "SOUTH" = '3', "WEST" = '4')

divisions <- list("NEW ENGLAND" = '1', "MIDDLE ATLANTIC" = '2', "EAST NORTH CENTRAL" = '3', "WEST NORTH CENTRAL" = '4', "SOUTH ATLANTIC" = '5', "EAST SOUTH CENTRAL" = '6', "WEST SOUTH CENTRAL" = '7', "MOUNTAIN" = '8', "PACIFIC" = '9')

# time code dictionaries
arrtime <- list('0' = NA, '1' = "12:02", '2' = "12:07", '3' = "12:12", '4' = "12:17", '5' = "12:22", '6' = "12:27", '7' = "12:35", '8' = "12:42", '9' = "12:47", "10" = "12:55", "11" = "1:02", "12" = "1:07", "13" = "1:12", "14" = "1:17", "15" = "1:22", "16" = "1:27", "17" = "1:32", "18" = "1:37", "19" = "1:42", "20" = "1:47", "21" = "1:55", "22" = "2:02", "23" = "2:07", "24" = "2:12", "25" = "2:17", "26" = "2:22", "27" = "2:27", "28" = "2:32", "29" = "2:37", "30" = "2:42", "31" = "2:47", "32" = "2:52", "33" = "2:57", "34" = "3:02", "35" = "3:07", "36" = "3:12", "37" = "3:17", "38" = "3:22", "39" = "3:27", "40" = "3:32", "41" = "3:37", "42" = "3:42", "43" = "3:47", "44" = "3:52", "45" = "3:57", "46" = "4:02", "47" = "4:07", "48" = "4:12", "49" = "4:17", "50" = "4:22", "51" = "4:27", "52" = "4:32", "53" = "4:37", "54" = "4:42", "55" = "4:47", "56" = "4:52", "57" = "4:57", "58" = "5:02", "59" = "5:07", "60" = "5:12", "61" = "5:17", "62" = "5:22", "63" = "5:27", "64" = "5:32", "65" = "5:37", "66" = "5:42", "67" = "5:47", "68" = "5:52", "69" = "5:57", "70" = "6:02", "71" = "6:07", "72" = "6:12", "73" = "6:17", "74" = "6:22", "75" = "6:27", "76" = "6:32", "77" = "6:37", "78" = "6:42", "79" = "6:47", "80" = "6:52", "81" = "6:57", "82" = "7:02", "83" = "7:07", "84" = "7:12", "85" = "7:17", "86" = "7:22", "87" = "7:27", "88" = "7:32", "89" = "7:37", "90" = "7:42", "91" = "7:47", "92" = "7:52", "93" = "7:57", "94" = "8:02", "95" = "8:07", "96" = "8:12", "97" = "8:17", "98" = "8:22", "99" = "8:27", "100" = "8:32", "101" = "8:37", "102" = "8:42", "103" = "8:47", "104" = "8:52", "105" = "8:57", "106" = "9:02", "107" = "9:07", "108" = "9:12", "109" = "9:17", "110" = "9:22", "111" = "9:27", "112" = "9:32", "113" = "9:37", "114" = "9:42", "115" = "9:47", "116" = "9:52", "117" = "9:57", "118" = "10:02", "119" = "10:07", "120" = "10:12", "121" = "10:17", "122" = "10:22", "123" = "10:27", "124" = "10:32", "125" = "10:37", "126" = "10:42", "127" = "10:47", "128" = "10:52", "129" = "10:57", "130" = "11:02", "131" = "11:07", "132" = "11:12", "133" = "11:17", "134" = "11:22", "135" = "11:27", "136" = "11:32", "137" = "11:37", "138" = "11:42", "139" = "11:47", "140" = "11:52", "141" = "11:57", "142" = "12:02", "143" = "12:07", "144" = "12:12", "145" = "12:17", "146" = "12:22", "147" = "12:27", "148" = "12:32", "149" = "12:37", "150" = "12:42", "151" = "12:47", "152" = "12:52", "153" = "12:57", "154" = "13:02", "155" = "13:07", "156" = "13:12", "157" = "13:17", "158" = "13:22", "159" = "13:27", "160" = "13:32", "161" = "13:37", "162" = "13:42", "163" = "13:47", "164" = "13:52", "165" = "13:57", "166" = "14:02", "167" = "14:07", "168" = "14:12", "169" = "14:17", "170" = "14:22", "171" = "14:27", "172" = "14:32", "173" = "14:37", "174" = "14:42", "175" = "14:47", "176" = "14:52", "177" = "14:57", "178" = "15:02", "179" = "15:07", "180" = "15:12", "181" = "15:17", "182" = "15:22", "183" = "15:27", "184" = "15:32", "185" = "15:37", "186" = "15:42", "187" = "15:47", "188" = "15:52", "189" = "15:57", "190" = "16:02", "191" = "16:07", "192" = "16:12", "193" = "16:17", "194" = "16:22", "195" = "16:27", "196" = "16:32", "197" = "16:37", "198" = "16:42", "199" = "16:47", "200" = "16:52", "201" = "16:57", "202" = "17:02", "203" = "17:07", "204" = "17:12", "205" = "17:17", "206" = "17:22", "207" = "17:27", "208" = "17:32", "209" = "17:37", "210" = "17:42", "211" = "17:47", "212" = "17:52", "213" = "17:57", "214" = "18:02", "215" = "18:07", "216" = "18:12", "217" = "18:17", "218" = "18:22", "219" = "18:27", "220" = "18:32", "221" = "18:37", "222" = "18:42", "223" = "18:47", "224" = "18:52", "225" = "18:57", "226" = "19:02", "227" = "19:07", "228" = "19:12", "229" = "19:17", "230" = "19:22", "231" = "19:27", "232" = "19:32", "233" = "19:37", "234" = "19:42", "235" = "19:47", "236" = "19:52", "237" = "19:57", "238" = "20:02", "239" = "20:07", "240" = "20:12", "241" = "20:17", "242" = "20:22", "243" = "20:27", "244" = "20:32", "245" = "20:37", "246" = "20:42", "247" = "20:47", "248" = "20:52", "249" = "20:57", "250" = "21:02", "251" = "21:07")
arrtime <- append(arrtime, list("252" = "21:12", "253" = "21:17", "254" = "21:22", "255" = "21:27", "256" = "21:32", "257" = "21:37", "258" = "21:42", "259" = "21:47", "260" = "21:52", "261" = "21:57", "262" = "22:02", "263" = "22:07", "264" = "22:12", "265" = "22:17", "266" = "22:22", "267" = "22:27", "268" = "22:32", "269" = "22:37", "270" = "22:42", "271" = "22:47", "272" = "22:52", "273" = "22:57", "274" = "23:02", "275" = "23:07", "276" = "23:12", "277" = "23:17", "278" = "23:22", "279" = "23:27", "280" = "23:32", "281" = "23:37", "282" = "23:42", "283" = "23:47", "284" = "23:52", "285" = "23:57"))

deptime <- list('0' = NA, '1' = "12:15", '2' = "12:45", '3' = "1:15", '4' = "1:35", '5' = "2:05", '6' = "2:45", '7' = "3:05", '8' = "3:15", '9' = "3:25", "10" = "3:35", "11" = "3:45", "12" = "3:55", "13" = "4:05", "14" = "4:15", "15" = "4:25", "16" = "4:35", "17" = "4:45", "18" = "4:55", "19" = "5:02", "20" = "5:07", "21" = "5:12", "22" = "5:17", "23" = "5:22", "24" = "5:27", "25" = "5:32", "26" = "5:37", "27" = "5:42", "28" = "5:47", "29" = "5:52", "30" = "5:57", "31" = "6:02", "32" = "6:07", "33" = "6:12", "34" = "6:17", "35" = "6:22", "36" = "6:27", "37" = "6:32", "38" = "6:37", "39" = "6:42", "40" = "6:47", "41" = "6:52", "42" = "6:57", "43" = "7:02", "44" = "7:07", "45" = "7:12", "46" = "7:17", "47" = "7:22", "48" = "7:27", "49" = "7:32", "50" = "7:37", "51" = "7:42", "52" = "7:47", "53" = "7:52", "54" = "7:57", "55" = "8:02", "56" = "8:07", "57" = "8:12", "58" = "8:17", "59" = "8:22", "60" = "8:27", "61" = "8:32", "62" = "8:37", "63" = "8:42", "64" = "8:47", "65" = "8:52", "66" = "8:57", "67" = "9:02", "68" = "9:07", "69" = "9:12", "70" = "9:17", "71" = "9:22", "72" = "9:27", "73" = "9:32", "74" = "9:37", "75" = "9:42", "76" = "9:47", "77" = "9:52", "78" = "9:57", "79" = "10:05", "80" = "10:15", "81" = "10:25", "82" = "10:35", "83" = "10:45", "84" = "10:55", "85" = "11:05", "86" = "11:15", "87" = "11:25", "88" = "11:35", "89" = "11:45", "90" = "11:55", "91" = "12:05", "92" = "12:15", "93" = "12:25", "94" = "12:35", "95" = "12:45", "96" = "12:55", "97" = "13:05", "98" = "13:15", "99" = "13:25", "100" = "13:35", "101" = "13:45", "102" = "13:55", "103" = "14:05", "104" = "14:15", "105" = "14:25", "106" = "14:35", "107" = "14:45", "108" = "14:55", "109" = "15:05", "110" = "15:15", "111" = "15:25", "112" = "15:35", "113" = "15:45", "114" = "15:55", "115" = "16:05", "116" = "16:15", "117" = "16:25", "118" = "16:35", "119" = "16:45", "120" = "16:55", "121" = "17:05", "122" = "17:15", "123" = "17:25", "124" = "17:35", "125" = "16:45", "126" = "17:55", "127" = "18:05", "128" = "18:15", "129" = "18:25", "130" = "18:35", "131" = "18:45", "132" = "18:55", "133" = "19:15", "134" = "19:45", "135" = "20:15", "136" = "20:45", "137" = "21:05", "138" = "21:15", "139" = "21:25", "140" = "21:35", "141" = "21:45", "142" = "21:55", "143" = "22:05", "144" = "22:15", "145" = "22:25", "146" = "22:35", "147" = "22:45", "148" = "22:55", "149" = "23:15", "150" = "23:45")


# year range validation
  if (year < 2010 || year == 2020 || year > 2024){
    print("Please enter a year between 2010 and 2024, inclusive; 2020 is also not available.")
    return()
  } else {
    census_string <- paste0("https://api.census.gov/data/", year, 
    "/acs/acs1/pums?get=")
  } # initializes census query with appropriate year

# numeric variables validation  
  if (length(numvar) < 2 || 
      any(!(numvar %in% c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP",
                             "JWMNP", "PWGTP")))){
    print("Please enter at least *two* numerical values from 'AGEP', 'GASP', 'GRPIP', 'JWAP', 'JWDP', 'JWMNP', or 'PWGTP'.")
    return()
  } else if ("PWGTP" %notin% numvar) {
    census_string <- paste0(census_string, "PWGTP,", str_c(numvar, collapse = ","))
  } else {
    census_string <- paste0(census_string, str_c(numvar, collapse = ","))
  } # appends numeric variables to end of census query
  
# categorical variables validation
  if (any(!(catvar %in% c("FER", "HHL", "SCH", "SEX")))){
    print("Please enter at least one categorical values from 'FER', 'HHL', 'SCH', or 'SEX' (default).")
    return()
  } else {
    census_string <- paste0(census_string, ",", str_c(catvar, collapse = ","))
  } # appends categorical variable(s) to end of census query

# geographical level validation  
  if (geolevel %notin% c("STATE", "REGION", "DIVISION")){
    print("Please enter a geographical level from 'STATE' (default), 'REGION', or 'DIVISION' and (optionally) furnish a specific one, otherwise all will be returned.")
    return()
  } else {
    census_string <- paste0(census_string, "&for=", tolower(geolevel))
  } # appends geographic level to end of census query

# subset level validation
  if (is.null(subset)){
    census_string <- paste0(census_string, ":*&SCHL=24&key=5f3b8018114ef7902b34be730306e5ebf9102875")
    census <- GET(census_string)
    census
  } else if (geolevel == "STATE" && subset %in% names(states)){
    census_string <- paste0(census_string, ":", states[subset], "&SCHL=24&key=5f3b8018114ef7902b34be730306e5ebf9102875")
  } else if (geolevel == "REGION" && subset %in% names(regions)){
    census_string <- paste0(census_string, ":", regions[subset], "&SCHL=24&key=5f3b8018114ef7902b34be730306e5ebf9102875")
  } else if (geolevel == "DIVISION" && subset %in% names(divisions)){
    census_string <- paste0(census_string, ":", divisions[subset], "&SCHL=24&key=5f3b8018114ef7902b34be730306e5ebf9102875")
  } else {
    print("Please enter an appropriate two-letter state (default) abbreviation, region, or division and designate appropriately in the `geolevel` input.")
    return()
  } # looks up and appends appropriate number to end of census query

# queries the server
  census <- GET(census_string)

# converts the queried data into a tibble
  census_tibble <- census$content |>
    rawToChar() |>
    fromJSON() |>
    as_tibble(.name_repair = "universal_quiet") |>
    row_to_names(row_number = 1) |>
    clean_names()

# numeric data cleaning
  if("AGEP" %in% numvar){
    census_tibble$agep <- as.numeric(census_tibble$agep)
    }
  if ("GASP" %in% numvar){
    census_tibble$gasp <- as.numeric(census_tibble$gasp)
    }
  if ("GRPIP" %in% numvar){
    census_tibble$grpip <- as.numeric(census_tibble$grpip)
    }
  if ("JWMNP" %in% numvar){
    census_tibble$jwmnp <- as.numeric(census_tibble$jwmnp)
    }
  census_tibble$pwgtp <- as.numeric(census_tibble$pwgtp)

# time data cleaning
  if ("JWAP" %in% numvar){
    for (i in 1:length(census_tibble$jwap)){
    census_tibble$jwap[i] <- arrtime[[census_tibble$jwap[i]]]
    }
    census_tibble$jwap <- hm(census_tibble$jwap, quiet = TRUE)
    }
  if ("JWDP" %in% numvar){
    for (i in 1:length(census_tibble$jwdp)){
    census_tibble$jwdp[i] <- deptime[[census_tibble$jwdp[i]]]
    }
    census_tibble$jwdp <- hm(census_tibble$jwdp, quiet = TRUE)
  }
  
# categorical data cleaning
  if("FER" %in% catvar){
    census_tibble$fer <- factor(census_tibble$fer,
                                levels = c('0', '1', '2'),
                                labels = c("N/A", "Yes", "No"))
  }
  if("HHL" %in% catvar){
    census_tibble$hhl <- factor(census_tibble$hhl,
                                levels = c('0', '1', '2', '3', '4', '5'),
                                labels = c("N/A", "English only", "Spanish",
                                           "Other Indo-European", "Asian and
                                           Pacific Island", "Other"))
  }
  if("SCH" %in% catvar){
    census_tibble$sch <- factor(census_tibble$sch,
                                levels = c('0', '1', '2', '3'),
                                labels = c("N/A",
                                           "No attendance in past 3 months",
                                           "Yes, public school or college",
                                           "Yes, private school, college, or
                                           home school"))
  }
  if("SEX" %in% catvar){
    census_tibble$sex <- factor(census_tibble$sex,
                                levels = c('1', '2'),
                                labels = c("Male", "Female"))
  }
  census_tibble$schl <- factor(census_tibble$schl,
                               levels = c('0', '1', '2', '3', '4',
                                          '5', '6', '7', '8', '9',
                                          '10', '11', '12', '13', '14',
                                          '15', '16', '17', '18', '19',
                                          '20', '21', '22', '23', '24'),
                               labels = c("N/A", "No schooling completed",
                                          "Nursery school, preschool",
                                          "Kindergarten", "Grade 1", "Grade 2",
                                          "Grade 3", "Grade 4", "Grade 5",
                                          "Grade 6", "Grade 7", "Grade 8",
                                          "Grade 9", "Grade 10", "Grade 11",
                                          "12th grade - no diploma",
                                          "Regular high school diploma",
                                          "GED or alternative credential",
                                          "Some college, less than 1 year",
                                          "1 or more years college, no degree",
                                          "Associate's degree", "Bachelor's
                                          degree", "Master's degree",
                                          "Professional degree beyond
                                          Bachelor's", "Doctorate degree"))
  
  census_tibble
}
```

